We Are All Trying to Describe Something And We Don't Have the Words

### 2023-10-31 outputs/paper/paper.qmd
RF section started, discussion too

TLDR of the Gamman interview is: Taste is subjective. People who can play instruments tend to be better at identifying and describing sounds. Cultural criticism is a hall of smoke and mirrors. Antonoff is everywhere.

Can we translate any of this into something observable? Can we quantify the Jack Antonoff-ification of pop music?

@@

It is a perfect video essay. I pestered him anyway:

@@

It's hard to gauge exactly what the variables "danceability" or "valence" mean, even using the documentation. For instance, "valence" is described as "A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry)". "Danceability" is described as "how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable". Neither of these definitions fully cover what metrics and thresholds go into something like "danceability". However, whatever these features capture is easier to compare across artists in visualizations than more granular details about bars, beats, duration, and tatums[^3].

[^3]: \-- a word I just learned, which refers to "the lowest regular pulse train that a listener intuitively infers from the timing of perceived musical events".

@@

I interviewed Glen McDonald, a Principal Engineer at Spotify who worked at The Echo Nest, a music intelligence start-up that was acquired by Spotify in 2014. Macdonald told me that `valence` and `danceability` were created by giving thousands of examples to college interns and asking them to tag whether a song was positive or gloomy, or danceable or un-danceable. Variables like energy or instrumentalness, on the other hand, were determined primarily through machine listening techniques. The human subjectivity part only came in when it came to tweaking the training data[^4]:

[^4]: My favourite example Glen gave was that the ML kept giving bluegrass songs insanely high `speechiness` ratings. Because there were no banjos in the training data, they were registered as human speech. (This is hilarious.) They had to go back and add more songs with banjos to the training data.

"You could imagine writing a formula for energy that combines like loudness and tempo and degree of harmonic variation or something. So it was [machine learning], but that one's more like a human helping a machine figure out a formula. Whereas valence is teaching the machine to try to reproduce a purely human thing. Same with danceability. I mean, danceability is whether a human can dance to it. So like, the machine's not gonna dance, so the machine can't have any opinion on that. The computer could have an opinion on energy. And the computer can definitely have an opinion on loudness. So there's a spectrum from, you know, loudness as purely analytical, and then energy is a little like loudness, with a little more subjectivity. And then danceability and valence are purely subjective."

@@

@@

@@

The Spotify Features also wouldn't help you understand in the same way you end up understanding it if you sat on my friend Matthew Jordan's couch and watched him beat box a few different Jack Antonoff and Pharrell impressions in order to show the contrast.

@@

This was a feature that was invented in a lab \~15 years ago. This number refers to the time a bunch of undergraduates were paid (hopefully?) to listen to a bunch of songs and tag whether they were danceable or not. It is pretty much based on intuition. It captures this je-ne-sais-quoi essence of music that connects with listeners on a visceral level. Even if it doesn't make YOUR body want to move, you could imagine it making SOMEBODY'S body want to move.

### In which I do a close listen of our model's top 5 most danceable songs[^5]:

[^5]: I encourage you to follow along and drop me an email with your own assessment at michaela.drouillard\@mail.utoronto.ca.

@@

    ![](images/Macefonian_phalanx.png)

@@

## Draw on the random forest danceability correlations graph, tie it into the close listen

## Danceability isn't even necessarily about Dancing! I propose: Visceral-ness

>>>>

@@

@@

It's hard to gauge exactly what the variables "danceability" or "valence" mean, even using the documentation. For instance, "valence" is described as "A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry)". "Danceability" is described as "how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable". Neither of these definitions fully cover what metrics and thresholds go into something like "danceability". However, whatever these features capture is easier to compare across artists in visualizations than more granular details about bars, beats, duration, and tatums[^1].

[^1]: \-- a word I just learned, which refers to "the lowest regular pulse train that a listener intuitively infers from the timing of perceived musical events".

@@

I interviewed Glen McDonald, a Principal Engineer at Spotify who worked at The Echo Nest, a music intelligence start-up that was acquired by Spotify in 2014. Macdonald told me that `valence` and `danceability` were created by giving thousands of examples to college interns and asking them to tag whether a song was positive or gloomy, or danceable or un-danceable. Variables like energy or instrumentalness, on the other hand, were determined primarily through machine listening techniques. The human subjectivity part only came in when it came to tweaking the training data[^2]:

[^2]: My favourite example Glen gave was that the ML kept giving bluegrass songs insanely high `speechiness` ratings. Because there were no banjos in the training data, they were registered as human speech. (This is hilarious.) They had to go back and add more songs with banjos to the training data.

"You could imagine writing a formula for energy that combines like loudness and tempo and degree of harmonic variation or something. So it was \[machine learning\], but that one's more like a human helping a machine figure out a formula. Whereas valence is teaching the machine to try to reproduce a purely human thing. Same with danceability. I mean, danceability is whether a human can dance to it. So like, the machine's not gonna dance, so the machine can't have any opinion on that. The computer could have an opinion on energy. And the computer can definitely have an opinion on loudness. So there's a spectrum from, you know, loudness as purely analytical, and then energy is a little like loudness, with a little more subjectivity. And then danceability and valence are purely subjective."

@@

Random forests (RF) build an ensemble of decision trees during the training phase, and then use majority voting for classification tasks. Each tree in the ensembled is trained on a random sample of data taken with replacement, and only a random subset of features are considered when splitting at each node. This method ensures that each tree in the ensemble is unique, avoiding overfitting and capturing non-linear relationships and more complex patterns in the data. The formula behind the RF is:

## Data Preparation

The categorical feature 'producer', which contains the names of the producer for each track, was one-hot encoded to create binary columns for each producer ('is_antonoff', 'is_elworth', etc.). The producer feature was then converted into a factor variable, and the dataset was split into training and tests sets.

## Model Training

The RF was trainied using the @ranger package. The 'importance' parameter was set to 'impurity', which sets the model to interpret variable importance using Gini impurity.

Gini impurity is a measure of misclassification. At every split in a decision tree, the algorithm tests how well each feature splits the data. The features that split the data the most accurately, and then with the least impurity, would have the lowest impurity score. In Random Forests, a variable's importance can be calculated by looking at how much the tree nodes that use that variable reduce impurity on average.

Gini impurity was chose on the grounds that this study is interested in not only classifying producers, but, more importantly, in understanding how the underlying data informs these decisions. By calculating the Gini impurity not only for the entire model, but specifically for each producer, we can better understand how the variables in our dataset describe the differences between producers. Without song structure, instruments, sound, or other more ephemeral components of style, these differences can be taken as a proxy for style.

## Results

### Model Performance

### Stylistic Difference

Correlation heatmaps can help us better understand the differences between each producer's "style". The variables with the most variance were danceability, loudness, and energy.

```{r}
#| echo: false
#| warning: false
## Range of correlation scores

library(dplyr)

range_data <- table_data %>%
  gather(variable, value, -producer, -term) %>%
  group_by(variable) %>%
  summarize(
    min_value = min(value, na.rm = TRUE),
    max_value = max(value, na.rm = TRUE),
    range = max_value - min_value
  ) %>%
  arrange(-range)

##Variables with greatest variance##
top_vars <- range_data %>% 
  top_n(3, range) %>%
  pull(variable)

print(top_vars)

selected_data <- table_data %>%
  select(producer, term, all_of(top_vars)) %>%
  arrange(producer, term)

print(selected_data)

library(ggplot2)
```

Similar to the logistic regression, danceability was also, across all producers, the most important predictor in the classification process.

```{r}
#| echo: false
#| warning: false
list_importance <- list()
unique_producers <- unique(data_reduced$producer)

for (producer in unique_producers) {
  data_temp <- data_reduced
  data_temp$producer_binary <- as.factor(ifelse(data_temp$producer == producer, 1, 0))

  rf_temp <- randomForest(producer_binary ~ . - producer, data = data_temp, ntree = 100)

  importance_temp <- rf_temp$importance
  print(colnames(importance_temp))  # To see available columns

  list_importance[[producer]] <- importance_temp[,'MeanDecreaseGini']
}

df_importance <- bind_rows(list_importance, .id = "producer")

head(df_importance)

df_importance_long <- df_importance %>%
  gather(variable, importance, -producer)

ggplot(df_importance_long, aes(x = reorder(variable, importance), y = importance)) +
  geom_bar(stat = "identity") +
  facet_wrap(~producer, scales = "fixed") + 
  coord_flip() +
  labs(title = "Variable Importance by Producer", 
       x = "Variable", 
       y = "Mean Decrease Gini") +
  theme_minimal()

In the visualization of danceability correlations across producers, Antonoff's danceability has the least positive correlation with valence (SCORE), the most negative correlation with acousticness (SCORE), and significant positive correlations with speechiness (SCORE), loudness (SCORE), and energy (SCORE).

```{r}
#| echo: false
#| warning: false
# Danceability Plot
ggplot(selected_data, aes(x=term, y=danceability, fill=producer)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="Danceability across Producers", y="Danceability Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

Antonoff's danceability score being, compared to other producers, significantly positively correlated with speechiness, loudness and energy, negatively correlated with acousticness, and less than average with valence tracks with how critics tend to describe his music.

His tracks have been described as anthemic (cite), with a distinct vocal treatment that makes it sound, as Gamman put it, "the way it would sound to the person who's singing them".

> In the treble mix of the song, like in the very upper end, sort of above where the vocals are -- he often really crushes that down. It's lower in volume than anything else, which is sort of a weird effect. When you hear someone speaking, you hear a lot of that, like, noise in their voice. When you yourself are speaking, you hear less of it, right? You hear more of your own voice bouncing around in your head. And so it sort of creates this effect of -- **like he mixes his vocals the way it would sound to the person who's singing , them. Which is sort of strange.** He often has no vocals on that upper end, which is very unusual, and then often there are random little bits of noise happening up there which is the sort of thing her likes to do.

Clearing noise and isolating the middle range of the vocals may make for a stronger "speechiness" feature.

Valence, like danceability, was also a subjective variable built using labeling. Returning to the music itself, Antonoff-produced songs with the highest danceability scores aren't exactly dance hits.

    -   Not danceable.

    -   Not danceable, but it does seem like it's trying to get your body to do something.

    -   More anthemic than danceable. It also seems like it's trying to coercing your attention, if not your body, and the speechiness, loudness, and energy are all apparent. The valence is apparent in the lyrics, but it doesn't have the kind of rhythm for dancing.

    -

@@

@@

### In which I do a close listen of our model's top 5 most danceable songs[^3]:

[^3]: I encourage you to follow along and drop me an email with your own assessment at michaela.drouillard\@mail.utoronto.ca.

@@

@@
